% front matter / titelei
% recto: halftitlepage / schmutztitelseite, vortitel
\thispagestyle{empty}
{
    \begin{fullwidth}
        \centering
        \hphantom{.}
        \vfill
        {\Huge
            Hunting Echoes\\
            for\\
            Auditory Scene Analysis
        }
        \vfill
        \vfill
    \end{fullwidth}
}

\clearpage{}

% verso: frontispiece / frontiszipzseite, vakatseite
% dedication
\cleardoublepage{}

% recto: titlepage / titelblatt, innentitel
\thispagestyle{empty}
{
    \calccentering{\unitlength}
    \begin{adjustwidth*}{\unitlength}{-\unitlength}
        \raggedleft{}
        {\Huge\color{Burgundy}%
        Hunting Echoes\\
        for Auditory Scene Analysis}\\[\baselineskip]
        {\LARGE%
        Dissertation Thesis}\\[0.2\textheight]
        {\huge%
        Diego \textsc{Di Carlo}}\\[\baselineskip]
        {\LARGE%
        \today}
        \vfill
        \vfill
        {\large%
        Submitted in partial fulfillment of the requirements\\
        for the degree of Doktor der Naturwissenschaften\\[\baselineskip]% (Dr.\ rer.\ nat.)\\[\baselineskip]

        to the\\[\baselineskip]

        Faculty of Mathematics\\
        at Ruhr-Universität Bochum\\[2\baselineskip]
        %Vorgelegt zur Erlangung des\\
        %Doktorgrades der Naturwissenschaften\\
        %an der Fakultät der Mathematik\\
        %der Ruhr-Universität Bochum\\[2\baselineskip]

        \begin{minipage}{0.5\textwidth}
        \begin{tabular}{lr}
            1st\hspace{4pt} Reviewer & Prof.\ Dr. Gregor Leander\\
            2nd Reviewer & Prof.\ Dr.\; Alexander May
        \end{tabular}
        \end{minipage}
        \hspace*{36pt}
        %}\hspace*{-8pt}

        %\vspace{2\baselineskip}
        %Datum der Disputation: 13.\ Dezember 2019
        \vfill
        }
        \vspace*{\baselineskip}
    \end{adjustwidth*}
}

\clearpage{}

% verso: colophon / impressum
\thispagestyle{empty}
\hphantom{.}
\vfill

\section*{Imprint}

\textit{Hunting Echoes for Audtory Scene Analysis}\\
Copyright \textcopyright{} 2020 by \theauthor{}.\\
All rights reserved. Printed in France.\\
Published by the Ruhr-Universität Bochum, Bochum, Germany.

\section*{Colophon}

This thesis was typeset using \LaTeX{} and the \texttt{memoir} documentclass.
It is based on Aaron Turon's thesis \emph{Understanding and expressing scalable concurrency}\footnote{\url{https://people.mpi-sws.org/~turon/turon-thesis.pdf}}, itself a mixture of \texttt{classicthesis}\footnote{\url{https://bitbucket.org/amiede/classicthesis/}} by Andr\'e Miede and \texttt{tufte-latex}\footnote{\url{https://github.com/Tufte-LaTeX/tufte-latex}}, based on Edward Tufte's \emph{Beautiful Evidence}.\\[0.5\baselineskip]
%
The bibliography was processed by Biblatex.
All graphics and plots are made with PGF/Ti\emph{k}Z.\\[0.5\baselineskip]
%
The body text is set 10/14pt (long primer) on a 26pc measure.
The margin text is set 8/9pt (brevier) on a 12pc measure.
Matthew Carter's \textrm{Charter} acts as both the text and display typeface.
Monospaced text uses Jim Lyles's \texttt{Bitstream Vera Mono} (\enquote{Bera Mono}).

\clearpage{}

% recto: dedication or epigraph
\thispagestyle{empty}
\vphantom{.}
\vfill
{%
    \flushright{}
    \emph{Pleasure to me is wonder—the unexplored, the unexpected, \\
    the thing that is hidden and the changeless thing\\
    that lurks behind superficial mutability.}\\
    \hfill--- Howard Phillips \textsc{Lovercraft}
}
\vfill
\vfill

% verso: blank
\clearpage{}
